
# loading libraries

```{r libraries}
library(edgeR)
library(compositions)
library(recount)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(cowplot)
library(readxl)
library(colorspace)
```

# loading glycogenes

```{r glycogene_ids}
glycogene_lookup=readRDS('glycogene_lookup.Rds')
```

# loading housekeeping genes

```{r housekeeping_ids}
seg_lookup = readRDS('seg_lookup.Rds')
```

# using DPAGT1 as reference gene in bulk transforms

```{r dpagt1_ref_gene}
# 7e8fb4c9b050a8d9e100d038629a0467a3a9c805
readTissueData_DPAGT1 = function(rse_gene,tissue,samples) {

  mtx = assay(rse_gene)[,samples]

  y = DGEList(counts=mtx,group=rep('X',ncol(mtx)))
  y = calcNormFactors(y)

  count_clr = apply(cpm(y),2,compositions::clr)
  rownames(count_clr) = stringr::str_replace(rownames(count_clr),'\\..*$','')
  glycogenes = count_clr[intersect(rownames(count_clr),glycogene_lookup),]
  reverse_glycogenes = setNames(names(glycogene_lookup),glycogene_lookup)
  rownames(glycogenes) = reverse_glycogenes[rownames(glycogenes)]
  genenames = rownames(glycogenes)

  dpagt_ref=glycogenes['DPAGT1',]

  stats = as.data.frame(t(apply(glycogenes, 1, function(row) {
    c(mean=mean(row),
      rel_mean=mean(row/dpagt_ref),
      rel_diff=mean(row-dpagt_ref),
      stdev=sd(row),
      iqr=iqr(row),
      samples=length(row))
  })))
  cbind(stats,tissue=tissue,gene=genenames)

}
```

# using each housekeeping gene in bulk transforms

```{r per_ref_gene}
# 4a3fee9b7b7d552846666cfbe113d124be7d6c5b
readTissueData_perHouseKeepingGene = function(rse_gene,tissue,samples) {

    mtx = assay(rse_gene)[,samples]

  # We want to normalise the count data per tissue set
  y = DGEList(counts=mtx,group=rep('X',ncol(mtx)))
  y = calcNormFactors(y)
  # Apply the clr to smooth out the data a bit
  count_clr = apply(cpm(y),2,compositions::clr)
  rownames(count_clr) = stringr::str_replace(rownames(count_clr),'\\..*$','')

  gene_lookup = c(glycogene_lookup, seg_lookup)
  genes = count_clr[intersect(rownames(count_clr), gene_lookup),]
  reverse_genes = setNames(names(gene_lookup), gene_lookup)
  rownames(genes) = reverse_genes[rownames(genes)]
  # We want to collect the mean clr, sd of the clr, IQR of the clr
  # This will give us the sets of fold-changes for genes that we expect to
  # see within normal tissue function -> Envelope for cell-level change

  ref = genes[intersect(rownames(genes), c('DPAGT1', names(seg_lookup))),]
  glycogenes <- genes[intersect(rownames(genes), names(glycogene_lookup)),]
  genenames = rownames(glycogenes)

  stats =
    mapply(
      function(ref_row, ref_gene) {
        data <-
          as.data.frame(t(apply(glycogenes, 1, function(row) {
            c(mean=mean(row),
              rel_mean=mean(row/ref_row),
              rel_diff=mean(row-ref_row),
              stdev=sd(row),
              iqr=iqr(row),
              samples=length(row))
            })))
        cbind(data, tissue = tissue, gene = genenames, ref_gene = ref_gene)
      },
      split(ref, rownames(ref)),
      names(split(ref, rownames(ref))),
      SIMPLIFY = F)

  stats = do.call(rbind, stats)

  stats

}
```

# using all housekeeping genes in bulk transform

```{r all_ref_genes}
readTissueData = function(rse_gene,tissue,samples) {

  mtx = assay(rse_gene)[,samples]

  # We want to normalise the count data per tissue set
  y = DGEList(counts=mtx,group=rep('X',ncol(mtx)))
  y = calcNormFactors(y)
  # Apply the clr to smooth out the data a bit
  count_clr = apply(cpm(y),2,clr)
  count_log = apply(cpm(y)+1,2,log)

  rownames(count_log) = stringr::str_replace(rownames(count_log),'\\..*$','')
  rownames(count_clr) = rownames(count_log)

  gene_lookup = c(glycogene_lookup, seg_lookup)
  genes = count_log[intersect(rownames(count_log), gene_lookup),]
  glycogenes_clr = count_clr[intersect((rownames(count_clr)),gene_lookup),]
  reverse_genes = setNames(names(gene_lookup), gene_lookup)
  rownames(genes) = reverse_genes[rownames(genes)]
  rownames(glycogenes_clr) = rownames(genes)

  glycogenes = genes[intersect(rownames(genes), names(gene_lookup)),]
  glycogenes_clr = glycogenes_clr[intersect(rownames(genes), names(gene_lookup)),]

  genenames = rownames(glycogenes)

  data = reshape2:::melt.matrix(glycogenes, varnames = c('gene','sample_id'), value.name = 'log_cpm', as.is = T)

  ref_data = data[data$gene %in% names(seg_lookup),]
  ref_data = aggregate.data.frame(
    x = list(seg_geom_avg = ref_data$log_cpm),
    by = list(sample_id = ref_data$sample_id),
    FUN = function(x) { exp(mean(log(x))) })

  data = merge(data, ref_data, by = 'sample_id')

  data$rel_frac = data$log_cpm/data$seg_geom_avg
  data$rel_diff = data$log_cpm-data$seg_geom_avg
  data$seg_geom_avg <- NULL

  stats = data.frame(
    stddev = apply(glycogenes, 1, sd),
    iqr = apply(glycogenes, 1, iqr),
    mean_clr = apply(glycogenes_clr, 1, mean),
    samples = ncol(glycogenes),
    gene = genenames,
    ref_gene = 'all')

  stats = merge(data, stats, by = c('gene'))
  stats$tissue = tissue

  stats
}
```

# loading recount2 tcga 

```{r load_tcga}
load('../../input/tcga/rse_gene_tcga.Rdata')

coldata = subset(colData(rse_gene),gdc_cases.samples.sample_type == 'Solid Tissue Normal')

# Group by tissue
all_tissues = unique(coldata[,'gdc_cases.project.primary_site'])
```

## tcga stats using dpagt1 as ref

```{r compute_tcga1}
tissue_results_tcga_DPAGT1 = lapply(all_tissues, function(tissue) {
  samples = rownames(subset(coldata, gdc_cases.project.primary_site == tissue))
  if (length(samples) > 1) {
    readTissueData_DPAGT1(rse_gene,tissue,samples)
  } else {
    return()
  }
})

tissue_results_tcga_DPAGT1 = do.call(rbind, tissue_results_tcga_DPAGT1)

saveRDS(tissue_results_tcga_DPAGT1,file = 'tissue_results_log_tcga_DPAGT1.Rds')
remove(tissue_results_tcga_DPAGT1)
```

## tcga stats using each housekeeping gene as ref

```{r compute_tcga2}
tissue_results_tcga_perHouseKeepingGene = lapply(all_tissues, function(tissue) {
  samples = rownames(subset(coldata, gdc_cases.project.primary_site == tissue))
  if (length(samples) > 1) {
    readTissueData_perHouseKeepingGene(rse_gene,tissue,samples)
  } else {
    return()
  }
})

tissue_results_tcga_perHouseKeepingGene = do.call(rbind, tissue_results_tcga_perHouseKeepingGene)

saveRDS(tissue_results_tcga_perHouseKeepingGene,file = 'tissue_results_log_tcga_perHouseKeepingGene.Rds')
remove(tissue_results_tcga_perHouseKeepingGene)
```

## tcga stats using all housekeeping gene as ref

```{r compute_tcga3}
tissue_results_tcga = lapply(all_tissues, function(tissue) {
  samples = rownames(subset(coldata, gdc_cases.project.primary_site == tissue))
  if (length(samples) > 1) {
    readTissueData(rse_gene,tissue,samples)
  } else {
    return()
  }
})

tissue_results_tcga = do.call(rbind, tissue_results_tcga)

saveRDS(tissue_results_tcga,file = 'tissue_results_log_tcga.Rds')
remove(tissue_results_tcga)
```

# removing tcga

```{r remove_tcga}
remove(rse_gene)
```

# loading recount2 gtex

```{r load_gtex}
load('../../input/gtex/rse_gene_gtex.Rdata')

# Group by tissue
all_tissues = unique(colData(rse_gene)[,'smts'])
```

## gtex stats using dpagt1 as ref

```{r compute_gtex1}
tissue_results_DPAGT1 = lapply(all_tissues, function(tissue) {
  message(tissue)
  samples = rownames(subset(colData(rse_gene),smts == tissue))
  readTissueData_DPAGT1(rse_gene,tissue,samples)
})

tissue_results_DPAGT1 = do.call(rbind, tissue_results_DPAGT1)

saveRDS(tissue_results_DPAGT1,file = 'tissue_results_log_DPAGT1.Rds')
remove(tissue_results_DPAGT1)
```

## gtex stats using each housekeeping gene as ref

```{r compute_gtex2}
tissue_results_perHouseKeepingGene = lapply(all_tissues, function(tissue) {
  message(tissue)
  samples = rownames(subset(colData(rse_gene),smts == tissue))
  readTissueData_perHouseKeepingGene(rse_gene,tissue,samples)
})

tissue_results_perHouseKeepingGene = do.call(rbind, tissue_results_perHouseKeepingGene)

saveRDS(tissue_results_perHouseKeepingGene,file = 'tissue_results_log_perHouseKeepingGene.Rds')
remove(tissue_results_perHouseKeepingGene)
```

## gtex stats using all housekeeping gene as ref

```{r compute_gtex3}
tissue_results = lapply(all_tissues, function(tissue) {
  message(tissue)
  samples = rownames(subset(colData(rse_gene),smts == tissue))
  readTissueData(rse_gene,tissue,samples)
})

tissue_results = do.call(rbind, tissue_results)

saveRDS(tissue_results,file = 'tissue_results_log.Rds')
remove(tissue_results)
```

# removing gtex

```{r remove_gtex}
remove(rse_gene)
```
