
```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)
library(ggpubr)
library(matrixStats)
library(cowplot)
```

```{r}
file.copy('../../preprocess/generate_bulk_data/seg_lookup.Rds', 'seg_lookup.Rds')
segs = unique(names(readRDS('seg_lookup.Rds')))
file.remove('seg_lookup.Rds')
```

```{r}
file.copy('../../preprocess/generate_bulk_data/tissue_results_log.Rds', 'tissue_results_log.Rds')
tissue_results = readRDS('tissue_results_log.Rds')
file.remove('tissue_results_log.Rds')
```

```{r}
file.copy('../../preprocess/generate_bulk_data/tissue_results_log_tcga.Rds', 'tissue_results_log_tcga.Rds')
tissue_results_tcga = readRDS('tissue_results_log_tcga.Rds')
file.remove('tissue_results_log_tcga.Rds')
```

```{r}
tissue_results_all = subset(
  rbind(
    cbind(tissue_results, source='gtex'),
    cbind(tissue_results_tcga, source='tcga')),
  !(gene %in% segs))
```

```{r}
file.copy('../../preprocess/generate_single_cell_data/pseudobulk_log_seg_norm.Rds', 'pseudobulk_log_seg_norm.Rds')
pseudobulk_rel_diff = readRDS('pseudobulk_log_seg_norm.Rds')
file.remove('pseudobulk_log_seg_norm.Rds')

cutoffs=attributes(pseudobulk_rel_diff)$cutoffs
median_cutoff= median(cutoffs)
pseudobulk_rel_diff$is_unknown=grepl('Unknown',pseudobulk_rel_diff$tissue)
pseudobulk_rel_diff=subset(pseudobulk_rel_diff, !(gene %in% segs))
```

# computing iqr, mean and median of SEG normalized counts for each gene in bulk and single cell datasets

Either using pseudobulks and bulk TPMs passing cutoff, or using all pseudobulks and all TPMs

```{r}
sc_data.df = bind_rows(
  filter(.data = pseudobulk_rel_diff, source == 'human' & rel_diff > median_cutoff) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff), num_cluster = n()) %>%
    mutate(cutoff = 'applied'),
  filter(.data = pseudobulk_rel_diff, source == 'human') %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff), num_cluster = n()) %>%
    mutate(cutoff = 'unapplied'))
```

```{r}
bulk_data.df = bind_rows(
  filter(.data = tissue_results_all, ref_gene == 'all' & samples > 10 & log_cpm > log(2)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'all', cutoff = 'applied'),
  filter(.data = tissue_results_all, ref_gene == 'all' & source == 'gtex' & samples > 10 & log_cpm > log(2)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'gtex', cutoff = 'applied'),
  filter(.data = tissue_results_all, ref_gene == 'all' & source == 'tcga' & samples > 10 & log_cpm > log(2)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'tcga', cutoff = 'applied'),
  filter(.data = tissue_results_all, ref_gene == 'all' & samples > 10 & log_cpm > log(1.05)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'all', cutoff = 'unapplied'),
  filter(.data = tissue_results_all, ref_gene == 'all' & source == 'gtex' & samples > 10 & log_cpm > log(1.05)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'gtex', cutoff = 'unapplied'),
  filter(.data = tissue_results_all, ref_gene == 'all' & source == 'tcga' & samples > 10 & log_cpm > log(1.05)) %>%
    select(gene, rel_diff) %>%
    group_by(gene) %>%
    summarize(`iqr(rel_diff)` = iqr(rel_diff), `median(rel_diff)` = median(rel_diff), `mean(rel_diff)` = mean(rel_diff)) %>%
    mutate(dataset = 'tcga', cutoff = 'unapplied'))
```

```{r}
plot_data.df = bind_rows(
    inner_join(filter(.data = sc_data.df, cutoff == 'applied'),   filter(.data = bulk_data.df, dataset == 'all'  & cutoff == 'applied'),   by = c('gene','cutoff')),
    inner_join(filter(.data = sc_data.df, cutoff == 'applied'),   filter(.data = bulk_data.df, dataset == 'gtex' & cutoff == 'applied'),   by = c('gene','cutoff')),
    inner_join(filter(.data = sc_data.df, cutoff == 'applied'),   filter(.data = bulk_data.df, dataset == 'tcga' & cutoff == 'applied'),   by = c('gene','cutoff')),
    inner_join(filter(.data = sc_data.df, cutoff == 'unapplied'), filter(.data = bulk_data.df, dataset == 'all'  & cutoff == 'unapplied'), by = c('gene','cutoff')),
    inner_join(filter(.data = sc_data.df, cutoff == 'unapplied'), filter(.data = bulk_data.df, dataset == 'gtex' & cutoff == 'unapplied'), by = c('gene','cutoff')),
    inner_join(filter(.data = sc_data.df, cutoff == 'unapplied'), filter(.data = bulk_data.df, dataset == 'tcga' & cutoff == 'unapplied'), by = c('gene','cutoff')))
```

```{r}
scatterplotter <- function(data.df, x.char, y.char, bulk_data.char, xlims, ylims) {

  cutoff_unapplied.df <- filter(.data = data.df, cutoff == 'unapplied' & dataset == bulk_data.char)
  cutoff_applied.df <- filter(.data = data.df, cutoff == 'applied' & dataset == bulk_data.char)

  # xlims = range(c(cutoff_unapplied.df[x.char], cutoff_applied.df[x.char]))
  # xlims = c(floor(xlims[1]), ceiling(xlims[2]))
  # ylims = range(c(cutoff_unapplied.df[y.char], cutoff_applied.df[y.char]))
  # ylims = c(floor(ylims[1]), ceiling(ylims[2]))

  ggplot_unapplied.obj <-
    ggplot() +
      geom_point(
        data = cutoff_unapplied.df,
        aes_string(x = as.name(x.char), y = as.name(y.char), color = 'num_cluster'), alpha = 0.5) +
      scale_color_gradientn(colours = colorRampPalette(c('red','orange'))(12)) +
      scale_x_continuous(breaks = seq(xlims[1], xlims[2], 1), limits = xlims) +
      scale_y_continuous(breaks = seq(ylims[1], ylims[2], 1), limits = ylims) +
      labs(color = 'all\nclusters') +
      theme_minimal() +
      theme(
        legend.position = 'right',
        axis.title.x.bottom = element_blank(),
        axis.title.y.left = element_blank())

  ggplot_applied.obj <-
    ggplot() +
      geom_point(
        data = cutoff_applied.df,
        aes_string(x = as.name(x.char), y = as.name(y.char), color = 'num_cluster')) +
      scale_color_gradientn(colours = colorRampPalette(c('darkblue','lightblue'))(12)) +
      scale_x_continuous(breaks = seq(xlims[1], xlims[2], 1), limits = xlims) +
      scale_y_continuous(breaks = seq(ylims[1], ylims[2], 1), limits = ylims) +
      labs(color = 'clusters\npassing\ncutoff') +
      theme_minimal() +
      theme(
        legend.position = 'right',
        axis.title.x.bottom = element_blank(),
        axis.title.y.left = element_blank())

  legend_unapplied <- ggpubr::get_legend(ggplot_unapplied.obj)
  legend_applied <- ggpubr::get_legend(ggplot_applied.obj)
  cor.obj <- cor.test(
    x = cutoff_applied.df[,c(x.char)][[1]],
    y = cutoff_applied.df[,c(y.char)][[1]],
    method = 'pearson',
    alternative = 'two.sided')

  aligned_plots <- align_plots(
    ggplot_unapplied.obj + theme(legend.position = 'none'),
    ggplot_applied.obj + theme(legend.position = 'none'),
    ggplot() +
      geom_text(
        data = cutoff_applied.df,
        aes(label = paste0('r = ', round(cor.obj$estimate, 2), '; p = ', signif(cor.obj$p.value, 3)), x = Inf, y = Inf),
        color = 'black', vjust = 'inward', hjust = 'inward') +
      geom_abline(slope = 1, intercept = 0) +
      scale_x_continuous(breaks = seq(xlims[1], xlims[2], 1), limits = xlims) +
      scale_y_continuous(breaks = seq(ylims[1], ylims[2], 1), limits = ylims) +
      xlab(paste('sc', gsub('x', '', x.char))) +
      ylab(paste('bulk', gsub('y', '', y.char))) +
      theme_minimal() +
      theme(legend.position = 'none'),
    align="hv", axis="tblr")

  ggplot.obj <- ggdraw(aligned_plots[[1]]) + draw_plot(aligned_plots[[2]]) + draw_plot(aligned_plots[[3]])

  result.list <- list(
    ggplot.obj,
    legend_unapplied,
    legend_applied)
}
```

# SupplementaryFigure3B

```{r, fig.height=7, fig.width=21}
ggpubr::annotate_figure(
  right = gridExtra::arrangeGrob(
    scatterplotter(data.df = plot_data.df, x.char = 'mean(rel_diff).x', y.char = 'mean(rel_diff).y', bulk_data.char = 'all',c(-10,0),c(-5,2))[[2]],
    scatterplotter(data.df = plot_data.df, x.char = 'mean(rel_diff).x', y.char = 'mean(rel_diff).y', bulk_data.char = 'all',c(-10,0),c(-5,2))[[3]],
    nrow = 2, ncol = 1),
  ggpubr::ggarrange(
    plotlist = list(
      scatterplotter(data.df = plot_data.df, x.char = 'mean(rel_diff).x', y.char = 'mean(rel_diff).y', bulk_data.char = 'all',c(-10,0),c(-5,2))[[1]],
      scatterplotter(data.df = plot_data.df, x.char = 'mean(rel_diff).x', y.char = 'mean(rel_diff).y', bulk_data.char = 'gtex',c(-10,0),c(-5,2))[[1]],
      scatterplotter(data.df = plot_data.df, x.char = 'mean(rel_diff).x', y.char = 'mean(rel_diff).y', bulk_data.char = 'tcga',c(-10,0),c(-5,2))[[1]]),
    ncol = 3, nrow = 1))
```

# SupplementaryFigure3C

```{r, fig.height=7, fig.width=21}
ggpubr::annotate_figure(
  right = gridExtra::arrangeGrob(
    scatterplotter(data.df = plot_data.df, x.char = 'iqr(rel_diff).x', y.char = 'iqr(rel_diff).y', bulk_data.char = 'all', c(0,5),c(0,4))[[2]],
    scatterplotter(data.df = plot_data.df, x.char = 'iqr(rel_diff).x', y.char = 'iqr(rel_diff).y', bulk_data.char = 'all', c(0,5),c(0,4))[[3]],
    nrow = 2, ncol = 1),
  ggpubr::ggarrange(
    plotlist = list(
      scatterplotter(data.df = plot_data.df, x.char = 'iqr(rel_diff).x', y.char = 'iqr(rel_diff).y', bulk_data.char = 'all',c(0,5),c(0,4))[[1]],
      scatterplotter(data.df = plot_data.df, x.char = 'iqr(rel_diff).x', y.char = 'iqr(rel_diff).y', bulk_data.char = 'gtex',c(0,5),c(0,4))[[1]],
      scatterplotter(data.df = plot_data.df, x.char = 'iqr(rel_diff).x', y.char = 'iqr(rel_diff).y', bulk_data.char = 'tcga',c(0,5),c(0,4))[[1]]),
    ncol = 3, nrow = 1))
```

# tossing genes into buckets delimited by bulk iqr order for each bulk dataset subset

```{r}
human_filtered_sc_bulk  <-
  filter(.data = plot_data.df, cutoff == 'applied' & dataset == 'all')  %>%
  arrange(`iqr(rel_diff).y`) %>%
  mutate(bulk_iqr = cut.default(`iqr(rel_diff).y`, c(seq(0, 2, by = 0.5), 4))) %>%
  mutate(xpos = as.numeric(bulk_iqr))
```

```{r}
human_filtered_sc_gtex <-
  filter(.data = plot_data.df, cutoff == 'applied' & dataset == 'gtex') %>%
  arrange(`iqr(rel_diff).y`) %>%
  mutate(bulk_iqr = cut.default(`iqr(rel_diff).y`, c(seq(0, 2, by = 0.5), 4))) %>%
  mutate(xpos = as.numeric(bulk_iqr))
```

```{r}
human_filtered_sc_tcga <-
  filter(.data = plot_data.df, cutoff == 'applied' & dataset == 'tcga') %>%
  arrange(`iqr(rel_diff).y`) %>%
  mutate(bulk_iqr = cut.default(`iqr(rel_diff).y`, c(seq(0, 2, by = 0.5), 4))) %>%
  mutate(xpos = as.numeric(bulk_iqr))
```

## computing pearson correlation between bulk and sc iqrs for each bulk dataset subset

```{r}
bulk_iqr_all_bulk.df  <-
  group_by(human_filtered_sc_bulk, bulk_iqr) %>%
  summarize(
    xpos = unique(xpos),
    cor = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$estimate,
    pvalue = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$p.value) %>%
  mutate(source = 'all')
```

```{r}
bulk_iqr_only_gtex.df <-
  group_by(human_filtered_sc_gtex, bulk_iqr) %>%
  summarize(
    xpos = unique(xpos),
    cor = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$estimate,
    pvalue = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$p.value) %>%
  mutate(source = 'gtex')
```

```{r}
bulk_iqr_only_tcga.df <-
  group_by(human_filtered_sc_tcga, bulk_iqr) %>%
  summarize(
    xpos = unique(xpos),
    cor = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$estimate,
    pvalue = cor.test(`iqr(rel_diff).x`, `iqr(rel_diff).y`, method = 'pearson', alternative = 'two.sided')$p.value) %>%
  mutate(source = 'tcga')
```

# SupplementaryFigure3D

```{r}
bulk_iqr_ordered.ggplot <-
  bind_rows(bulk_iqr_all_bulk.df, bulk_iqr_only_gtex.df, bulk_iqr_only_tcga.df) %>%
  ggplot(aes(x = xpos, y = cor, color = source)) +
  geom_point() +
  scale_x_continuous(
    breaks = c(0.5,1.5,2.5,3.5,4.5,5.5,6.5),
    labels = c(0,0.5,1,1.5,2,2.5,3)) +
  coord_cartesian(x = c(0.5,6.5), y = c(-0.75, 0.5)) +
  geom_smooth(method = 'loess', level = 0.5) +
  geom_label_repel(aes(label = signif(pvalue, 3)), color = 'black', size = 2) +
  xlab('bulk iqr delmited interval') +
  ylab('cor(iqr(BULK_rel_diff),iqr(SC_rel_diff))')

bulk_iqr_ordered.ggplot
```
